---
title: "Pre-course installation requirements"
author: "Dale L. Forrister"
affiliation: "School of Biological Sciences, University of Utah"
date: "9/16/2022"
contact: "dlforrister@gmail.com"
output: html_document
---

# 1 A quick note about R Notebooks:

-   All scripts have been written as R Notebooks. This allows us to have extensive annotation and description of code, including inserting images and tables. However to fully visualize these it's important to change in to "visual mode"

    -   On the tool bar above click back an forth between "source" and "visual".

-   There are two ways to run code within the notebook.

    -   As a SINGLE line. Click a on a line of code within a notebook and hit control + enter.

    -   the entire "chunk". If click the green arrow head on the top right of the chunk you will run the entire chunk, line by line

        TRY BOTH BELLOW

```{r}
print("Just a single line")

print("All of the lines")
```

# 2 Check your version of R

Most packages used in metabolomics are maintained via Bioconductor, and thus need to be installed with "BiocManager"

The analysis in this document requires an R version = 4.2.1 and recent versions of the MSnbase and xcms (version \>= 3.19 is needed) packages. The packages can be installed with:

First, check that R is the most recent verison

```{r}
R.version


```

If this is not 4.2.1 or later please update R here:

<https://cran.r-project.org/>

Download the correct version for your computer.

Restart Rstudio and check run the chunk above to confirm it's up to date.

# 3 Install the required packages

```{r packages, include=FALSE}

if (!requireNamespace("BiocManager", quietly=TRUE)) { 
    install.packages("BiocManager")}


if (!requireNamespace("xcms", quietly=TRUE)) {
    BiocManager::install(version='devel')
    BiocManager::install("xcms")}

library(xcms)

```

Again, let's confirm it is the most recent version (3.19)

```{r}
package.version("xcms")
```

In this tutorial we will also use the here package in order to simplify file paths when working with R projects and R markdown files.

```{r filepaths}
if (!require(dplyr)) {install.packages('dplyr')}

library(dplyr)

if (!require(here)) {install.packages('here')}

library(here)
here()
```

The here library is very usefull when working with R projects. Using here() provides a short hand notation that is always based on where the .rproj file is located.

When we run here() it will tell us our starting point

then we can build file paths relative to this starting point using "," between folders followed by a file path name.

```{r}
here("data","standards","RN18_50_222.mzML")

```
